---
title: "Reproducible Research: Peer Assessment 1"
author: "Joaquín Soliño Pérez"
date: "Jan, 17th, 2021"
output: 
  html_document:
    keep_md: true
---

## Intro

This documment includes the work I have made for the Assesstment 1 of the course "Reproducible Research".

More than try to deliver the assigment, I've tried to test things.
I have wrote this as I were explaining the procedure step-by-step.

Sorry for the long explanation. I hope you fid it clear. Regards

## Loading and preprocessing the data

In first place we have to load the libraries we'll use.

```{r LOAD_&_PREP_0, echo=TRUE}
library(lubridate)      # lubridate for date management
library(lattice)        # lubridate plot system
library(dplyr)          # for data transformation - left_join
library(tidyr)          # for data transformation - pivot_longer

```

### Loading data (Q1)
Next, you unzip the downloaded file `activity.zip`, read the content and store it in a new data frame named `act_data`. This is the base date wich I will work all the assesstment.

```{r LOAD_&_PREP_1, echo=TRUE}
unzip("activity.zip")                # unzip downloaded file
act_data <- read.csv("activity.csv") # assign data to act_data data frame 

```

We take a look to `act_data`:

```{r LOAD_&_PREP_2, echo=TRUE}
dim(act_data) 

```

We have a data frame with 17568 rows and 3 columns (variables).
Let's inspect the data frame:

```{r LOAD_&_PREP_3, echo=TRUE}
head(act_data)             # first rows

tail(act_data)             # last rows

```

and finally we use `str` to see the structure:

```{r LOAD_&_PREP_4, echo=TRUE}
str(act_data) 

```

You can see 3 variables: 

- `steps`: an integer with number os recorded steps (and with NA records)
- `date`: a string in format "YYYY-MM-DD"
- `interval`: an integer that represents the hour and minutes of the record

Lets count the number of `NAs`:

```{r LOAD_&_PREP_5, echo=TRUE}
sum(is.na(act_data$steps))    # NAs in steps variable
sum(is.na(act_data$date))    # NAs in data variable
sum(is.na(act_data$interval))    # NAs in interval variable

```

> We find 2.304 / 17.568 x 100% = 13,05% of `NAs` in the `steps`  variable. There is no `NAs`for date and/or interval variables.

### Prepare de data

To answer the questions, we'll need to make some transformations:

- Convert `date` to a date-time
- Create a new variable `wday`, with the number of weekday from `date`.
- Create a new variable `tday`, wich distinguish weekdays and weekends. This variable is going to be a ordered factor with two levels: "weekday" and "weekend".

Remember that I use the "lubridate" library for date/time management.
Next, the transformations:

```{r LOAD_&_PREP_7, echo=TRUE}
# Convert string to date with ymd() function
act_data$date <- ymd(act_data$date)

# new variable wday (day of the week) - I avoid labels due to different language
# and week start criteria in different countries
act_data$wday <- factor(
      wday(act_data$date, week_start = 1),      # 1- start on mondays
      levels = c(1, 2, 3, 4, 5, 6, 7),
      ordered = TRUE
)

# tday - type of day - weekday or weekend
act_data$tday[act_data$wday<=5] <- "weekday"    # 1 - 5 mon to fri
act_data$tday[act_data$wday>=6] <- "weekend"    # 6 - 7 sat and sun

# Convert tday in a factor with levels "weekday" and "weekend"
act_data$tday <- factor(act_data$tday,
                        levels = c("weekday", "weekend"),
                        ordered = TRUE)

```

Finally we have our data frame `act_data` ready for the analysis. Take a look to the structure:

```{r LOAD_&_PREP_8, echo=TRUE}
str(act_data) 

```

## What is mean total number of steps taken per day?

To calculate the mean of steps taken per day, first we need to summary the total steps per day, because our data has steps per interval in a day.

We'll use the function `aggregate` with our data frame `act_data`, grouping by `date` variable. We'll store the result in a new data frame named `steps_day`.

```{r MEAN_BY_DAY1, echo=TRUE}
steps_day <- aggregate(
      act_data$steps,
      by = list(act_data$date),           # aggregate on date
      FUN = function(x)                   # we use the function to avoid NAs
            sum(x, na.rm = TRUE)          
)
names(steps_day) <- c("date","sum")      # assign names to the result variables

str(steps_day)

```

`steps_day` is a data frame with two variables:

- `date`: (date) the date from the original data
- `sum`: (int) the sum of the steps fo all intervals for de date in `date`

### Mean and median (Q3)

Now, we can easilly compute mean and median over the variable `sum` of the data frame `steps_day`:

```{r MEAN_BY_DAY3, echo=TRUE}
mean_day <- mean(steps_day$sum,na.rm = TRUE)          # mean of total steps per day
median_day <- median(steps_day$sum,na.rm = TRUE)        # median of total steps per day

mean_day
median_day

```

### Histogram - total steps per day (Q2)

To understand better the pattern of day activity, we plot the histogram of total steps by day:

```{r MEAN_BY_DAY2, echo=TRUE}

histogram(date ~ sum,
      data = steps_day,
      breaks = seq(from = 0, to = 25000, by = 2500),
      type = "count",
      col = "wheat1",
      main = "Histogram - Total Steps per day",
      xlab = "Total Steps by day",
      ylab = "Frequency (# days)"
      
)
```


## What is the average daily activity pattern?

Now we go deep in understand daily activity pattern.
First, from `act_data`, we compute the mean of steps by interval over all the days (excluding `NAs`). We store the result in a data frame named `steps_interval`.

### Aggregate data by day

```{r MEAN_BY_INTERVAL1, echo=TRUE}
steps_interval <- aggregate(           # aggregates by interval all the days
      act_data$steps,
      by = list(act_data$interval),
      FUN = function(x)
            mean(x, na.rm = TRUE)
)
names(steps_interval) <- c("interval", "mean")  # assign names to variables
```

`steps_day` is a df wit two variables:

- `interval`: the number that identifies the time of the record
- `mean`: the mean of steps per inteval over all the dates

### Plot average steps per interval (Q4 & Q5)

Next we plot the time series of the 5min-intervals

```{r MEAN_BY_INTERVAL2, echo=TRUE}
# plot time series (type = "l")
xyplot(mean ~ interval, type = "l",
       data = steps_interval,
       main = "Daily pattern - average steps by interval (5 min)",
       xlab = "Interval",
       ylab = "# steps (mean)"
)

```

Lets see the summary of this pattern:
```{r MEAN_BY_INTERVAL3, echo=TRUE}
summary(steps_interval$mean)
```

We can see a the average peak before 1000. We want to locate the interval where de average steps are the maximum: 

```{r MEAN_BY_INTERVAL4, echo=TRUE}
max_interval <-
      steps_interval$interval[which.max(steps_interval$mean)]     # returns the interval where the mean steps are the maximum
max_interval

```

> The maximum of average steps in an interval is 206.17 at 8:30.

## Imputing missing values (Q6)

As we find in the beggining, we find 3.204 NAs in steps variable.

```{r MISSING_VALUES_1, echo=TRUE}
sum(is.na(act_data$steps))    # NAs in steps variable
```

We have several options to define a strategy to fill `NAs`. 
In general, is commonly acepted that the activity pattern depends on the day of the week.

>The strategy I propose is to fill each `NA` with the average number of steps for the same interval and the same day of the week (i.e. the `NA` of an interval 1430 in monday _wday = 1_ is filled with the average of interval 1430 of all the mondays.

Previous to make this procedure, we are going to test that this is a reasonable option with the data we have.

### Daily pattern by day of the week
First, we can check if there is any pattern over the day of the week, so we compute de average total steps per day of the week:

```{r MISSING_VALUES_3, echo=TRUE}
# 1st aggregate data by date an day of the week to sum the total steps per day
steps_day <- aggregate(
      act_data$steps,
      by = list(act_data$date,act_data$wday),  # aggregate by date and day of the week 
      FUN = function(x)
            sum(x, na.rm = TRUE)
)
names(steps_day) <- c("date", "wday", "sum")

# 2nd aggregate data by day of the week to compute the average total steps per day of the week
aggr_steps_day <- aggregate(
      steps_day$sum,
      by = list(steps_day$wday),
      FUN = function(x)
            mean(x, na.rm = TRUE)
)
names(aggr_steps_day) <- c("wday", "mean")
aggr_steps_day

```

We can see there is significant differences between every day of the week. We make a Plot to see it better:

```{r MISSING_VALUES_4, echo=TRUE}

barchart(mean ~ wday, type = "l",
       data = aggr_steps_day,
       col = "wheat1",
       main = "Average total steps by day of week",
       xlab = "Day of week (1-mon - 7-sun",
       ylab = "# Steps"
)

```

### Pattern of NAs ofer intervals and day of week

Next, we look for the distribution of NAs per interval and day of the week.

```{r MISSING_VALUES_6, echo=TRUE}

count_samples <- with(act_data,table(interval,wday))  # wday -> 1 monday - 7 sunday
head(count_samples)           # first lines
tail(count_samples)           # last lines
```

As you can see, there are 9 records per interval from Monday to Friday (1 to 5). And 8 records per interval for saturday an Sunday (6 and 7)

Following a similar procedure, we obtain the distribution of `NAs` per interval and day of the week:

```{r MISSING_VALUES_7, echo=TRUE}
count_nas <- with(act_data[is.na(act_data$steps),],table(interval,wday,exclude = NA))  # wday -> 1 monday - 7 sunday
head(count_nas)           # first lines
tail(count_nas)           # last lines
```

You can see that whe have, 2 `NAs` records per interval on Mondays (wday == 1) and Fridays (wday == 5). There is no NAs on Tuesdays (wday = 2), and there is 1 `NA` record per interval for the rest of the days of the week.

Finally, we comopute the ratio of NAs vs total of records.

```{r MISSING_VALUES_8, echo=TRUE}
ratio_nas <- count_nas / count_samples  # wday -> 1 monday - 7 sunday
head(ratio_nas)           # first lines
tail(ratio_nas)           # last lines

```

With these results, we can conclude the proposal we made at the beggining of this subject, is quite reasonable.


### Defining a strategy to fill NAs

Related to the analisys we have made previously, our strategy is to fill `NAs` of an specific interval for a specific date with the average number of steps for the same intefval and the same day of the week of de date we want to fill.

The steps to do it are:

1. Copy the data to a new data set, named `act_data_fill`, where we fill the NAs 
2. Summarise the data a by: 1. interval, 2. day of the week, and compute the mean of the steps.
3. Build a vector with the values to filled (for this, we'll use the left_join )

#### Build data frame with filled data

Create a copy of the `act_data`, called `act_data_fill`:

```{r FILL_NAS_1, echo=TRUE}
act_data_fill <- act_data

```

Obtain the mean number of steps per interval and day of the week, and store it in a variable `avg_steps_inter_wday`. 

#### Calculate average steps per interval and day of week

Compute the aggregate the mean of `act_data` by `wday` and `interval `variables. The result is assigned to a data frame named `avg_steps_inter_wday`.

```{r FILL_NAS_2, echo=TRUE}

avg_steps_inter_wday <-
      with(act_data, aggregate( steps, by = list(wday, interval),
                  FUN = function(x) round(mean(x, na.rm = TRUE),0)     # mean rounded
                  )
      )
names(avg_steps_inter_wday) <- c("wday","interval","mean")
head(avg_steps_inter_wday)          # Show data frame
```

> Note: I have rounded the mean to 0 decimal digits, because we are going to fill number of steps, and steps are integer.

Now we build a dataframe named as `fill_df`, with all the values to be filled. We'll use the `left_join` function from `dplyr` to paste the appropiate value in each record by `wday` and `interval`.

```{r FILL_NAS_3, echo=TRUE}

fill_df <- act_data %>% 
      filter(is.na(steps)) %>%            # fill vector only has the rows with NAs
      select(date, interval,wday) %>%     # we select, date, interval and wday - day of the week
      # we use left_join to preserve all the rows form act_data and join by interval and wday
      left_join(avg_steps_inter_wday, by = c("interval","wday"))

head(fill_df, 10)       # 10 first rows
```

Finally we fill all the `NAs` in the `filled` from `act_data_fill` with the values from `fill_df$steps`.

```{r FILL_NAS_4, echo=TRUE}
# duplicate steps variable in a new filled variable.
act_data_fill$filled <- act_data_fill$steps

# assign average steps stored in fill_df to the subset of NA of filled variable from act_data_fill
act_data_fill$filled[is.na(act_data_fill$steps)] <- fill_df$mean

act_data_fill <- act_data_fill %>% rename(registered = steps)

str(act_data_fill)

# check the number of NAs on filled data.
num_nas <-sum(act_data_fill$filled[is.na(act_data_fill$filled)])
num_nas
```

As you can see, there is no `NAs` in filled data.

#### Analyzing filled data (Q7)

Following the same process as before, we aggregate `act_data_fill` by `date` variable.

```{r FILL_NAS_5, echo=TRUE}
steps_day_fill <- aggregate(                   # aggregate on date
      act_data_fill$filled,
      by = list(act_data_fill$date),
      FUN = function(x)                   # we use the function to avoid NAs
            sum(x, na.rm = TRUE)          
)
names(steps_day_fill) <- c("date", "sum")      # assign names to the result variables

# Mean of filled variable
mean_day_fill <- mean(steps_day_fill$sum, na.rm = TRUE)
mean_day_fill
# Mean of original variable
mean_day

#How much bigger is the mean with filled data?
mean_day_fill / mean_day

```

> As you can see, the mean of filled data is 15.7% greater than the original.

We do the same with the median.

```{r FILL_NAS_6, echo=TRUE}
# Mean of filled variable
median_day_fill <- median(steps_day_fill$sum, na.rm = TRUE)
median_day_fill
# Mean of original variable
median_day

#How much bigger is the mean with filled data?
median_day_fill / median_day

```

> The median for filled data is almost 6% greater than the original data. 

To better understand of the reason os these changes, I make a histogram plot comparing registered (original) data vs filled data:

```{r FILL_NAS_7, echo=TRUE}
# Create a df with both registered and filled data pivoted with a new variable 
# origin, with two levels: "registered" or "filled".
act_data_union <-
      act_data_fill %>% pivot_longer(c("registered", "filled"),
      names_to = "origin",
      values_to = "steps")

# Convert Origin in a ordered factor
act_data_union$origin <-
      factor(
            act_data_union$origin,
            levels = c("registered", "filled"),
            ordered = TRUE
      )

# Aggregate the total steps per date, type of day and oigin
steps_day_union <- aggregate(
      act_data_union$steps,
      by = list(act_data_union$date, act_data_union$tday, act_data_union$origin),
      FUN = function(x)
            sum(x, na.rm = TRUE)
)
names(steps_day_union) <- c("date", "tday","origin","sum")

# plot the histogram using Lattice Plot systems, where whe plot side-by-side
# histogram with registered data (the same as above) and the histogram with 
# the filled data
histogram( ~ sum | origin, 
        data = steps_day_union,
        breaks = seq(from = 0, to = 25000, by = 2500),
        type = "count",
        col = "wheat1",
        main = "Distribuction Steps by day",
        xlab = "Steps per day",
        ylab = "Freq (# of days)"
)
```

You can see clearly the most difference between filled data and registered data:

* Significant decrease of number of days with less than 2500 steps (almost 12 to 3)
* Significant increase of number of days between 10000 an 12500 steps (almost 18 to more than 20)
* Increase o number of days between 7500 and 10000 steps (from 7 to 10)


## Are there differences in activity patterns between weekdays and weekends? (Q8)

To evaluate differences in daily patterns between weekdays and weekends we start with a summary for each one.

### Daily interval pattern

Firs, we aggregate the mean steps by interval and type of day. We repeat the process at the beggining with `steps_interval`, but including the `tday` variable:

```{r PATTERN_TYPE_DAY_1, echo=TRUE}
# Aggregate steps by interval and type of day - original data
steps_interval <- aggregate(          
      act_data$steps,
      by = list(act_data$interval, act_data$tday),
      FUN = function(x)
            mean(x, na.rm = TRUE)
)
names(steps_interval) <- c("interval","tday","mean")      # assign names to the result variables
```

You now summarizes each data by separate (weekdays and weekends):

```{r PATTERN_TYPE_DAY_2, echo=TRUE}

summary(steps_interval$mean[steps_interval$tday == "weekday"])

summary(steps_interval$mean[steps_interval$tday == "weekend"])

```

> You find mean and median in weekend is greater than weekday. Also you find than max average steps interval in weekday is higher that weekend ones.

Lets look in a time plot comparing weekend and weekday:

```{r PATTERN_TYPE_DAY_3, echo=TRUE}

xyplot(mean ~ interval | tday , type = "l",
       data = steps_interval,
       layout = c(1,2),
       main = "Daily pattern - average steps by interval (5 min)",
       xlab = "Interval",
       ylab = "# steps (mean)"
)

```

This view complements the summaries we compute just before.

Lets Compare daily interval average per type of day (weekday vs weekend) and from registered data vs filled data.

```{r PATTERN_TYPE_DAY_4, echo=TRUE}

# Aggregate the total steps per interval, type of day and oigin (registered and filled data)

steps_interval_union <- aggregate(
      act_data_union$steps,
      by = list(act_data_union$interval, act_data_union$tday, act_data_union$origin),
      FUN = function(x)
            mean(x, na.rm = TRUE)
)
names(steps_interval_union) <- c("interval", "tday","origin","mean")


xyplot(mean ~ interval | origin + tday , type = "l",
       data = steps_interval_union,
       layout = c(2,2),
        main = "TSeries - Steps by interval (5 min)",
        xlab = "Time",
        ylab = "# Steps"
        )
```

> There is no a clear difference between daily average steps per interval when we compare registered (original) data vs filled data. This makes sense, because we used the meaan steps per interval and day of the week to fill the NAs in each interval and day of the week.    

### Daily pattern

Lets compare differences between weekday and weekend, looking for the total number of steps per day

```{r PATTERN_TYPE_DAY_5, echo=TRUE}
# Aggregate the total steps per date, type of day and oigin
mean_day_union <- aggregate(
      steps_day_union$sum,
      by = list(steps_day_union$tday, steps_day_union$origin),
      FUN = function(x)
            mean(x, na.rm = TRUE)
)
names(mean_day_union) <- c("tday","origin","mean")
mean_day_union

```

Median
```{r PATTERN_TYPE_DAY_6, echo=TRUE}
# Aggregate the total steps per date, type of day and oigin
median_day_union <- aggregate(
      steps_day_union$sum,
      by = list(steps_day_union$tday, steps_day_union$origin),
      FUN = function(x)
            median(x, na.rm = TRUE)
)
names(median_day_union) <- c("tday","origin","median")
median_day_union

```




```{r PATTERN_TYPE_DAY_7, echo=TRUE}

# plot the histogram using Lattice Plot systems, where whe plot side-by-side
# histogram with registered data (the same as above) and the histogram with 
# the filled data
histogram( ~ sum | origin + tday, 
        data = steps_day_union,
        breaks = seq(from = 0, to = 25000, by = 2500),
        type = "count",
        col = "wheat1",
        main = "Compare Distribution of total steps per day by type of day (weekday vs weekend and registered vs filled)",
        xlab = "Steps per day",
        ylab = "Freq (# of days)"
)
```

We can take the next conclusions:

* In the weekend there 